preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-03 Manuscript/manuscript.Rmd
minor edit: combined two paragraphs
removed reference names
updated with full results for fig5

afgan2018galaxy, ramirez2016deeptools2

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix and plotHeatmap tools offered by DeepTools [@ramirez2016deeptools2] via integration with the Galaxy project [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test.

Furthermore, we assessed the hypothesis that TADs and chromatin loops are conserved  [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa] by comparing predicted and called boundaries between the GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

<!--
At this point, we opted to implement preciseTAD using an optimally built whole genome random forest model. That is, a random forest model built on 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143). The reason for this was because of the improved performance that was exhibited for a whole genome random forest compared to aggregated chromosome-specific models (Supplementary Figure 5). When tuning the full genome RF we found that performances converged when *mtry*=2, *ntree*=250, and *nodesize*=??? (Supplementary Figures 6-7). 
-->

@@

![**preciseTAD boundaries are enriched in CTCF sites.** (A) Visualizing locations of predicted vs called ARROWHEAD TAD boundaries on GM12878 (chr 11: 33000000-34600000). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. The blue vertical lines represent boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by our random forest. The green vertical lines represent the *preciseTAD boundary points* (PTBP) predicted by *preciseTAD*, with the corresponding demarcated TADs outlined above in green. Below is a zoomed in portion of the genome showing a cluster of base pairs with high probability representing a single *preciseTAD boundary region (PTBR) colored in yellow. (B) Enriched heatmaps comparing the enrichment of CTCF sites around 5 kb flanked preciseTAD and ARROWHEAD TAD boundaries respectively. (C) Enriched heatmaps comparing the enrichment of CTCF sites around 10 kb flanked preciseTAD and PEAKACHU loop boundaries respectively. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF site between preciseTAD boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively. (E) A atrix of pairwise Jaccard indices across resolutions shows more invariance toward resolution for *preciseTAD* (upper triangular) compared to ARROWHEAD (lower triangular).???update when ready](Figures/figure5_enrichment_and_log2dist_ctcf.png)

@@

**Supplementary Figure 5. Whole genome random forest model outperform chromosome-specific random forest models across resolutions.** Random forest models built on ARROWHEAD boundaries using 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143) applied to the entire genome (A) exhibit higher balanced accuracies than their chromosome specific counterparts (B) across each resolution respectively.

**Supplementary Figure 6. Whole genome RF performances converge when ntree=250.** Scatter plot of cross-validated balanced accuracies (BA) for random forests built using both ARROWHEAD (circles) and PEAKACHU (triangles) boundaries, with random under-sampling, distance-type predictors (CTCF, RAD21, SMC3, and ZNF143), over the entire binned genome, for GM12878 (red) and K562 (blue) respectively. For the ARROWHEAD model, 5kb bins were used, while 10kb bins were used for the PEAKACHU model. Performances converged after trees of size 250, marked by the dashed line.

**Supplementary Figure 7. Whole genome RF performances converge when nodesize=???.** Scatter plot of cross-validated balanced accuracies (BA) for random forests built using both ARROWHEAD (circles) and PEAKACHU (triangles) boundaries, with random under-sampling, distance-type predictors (CTCF, RAD21, SMC3, and ZNF143), over the entire binned genome, for GM12878 (red) and K562 (blue) respectively. For the ARROWHEAD model, 5kb bins were used, while 10kb bins were used for the PEAKACHU model. Performances converged after nodesize values of ???, marked by the dashed line.

@@

**Supplementary Table 6.**

**Supplementary Table 7.**

<!--
**Supplementary Figure 3. The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS).

**Supplementary Figure 5. Venn diagrams illustrating the amount of overlap between predicted and called TAD boundary regions.** For each of the GM12878 (A) and K562 (B), both predicted and called TAD boundaries were flanked other either side by 10 kb and the amount of overlap was enumerated.

**Supplementary Figure 6. Predicted TAD boundaries were centered closer to peak signal regions defined by known architectural transcription factors when compared to called TAD boundaries for K562.** Enrichment
heatmaps illustrate the average signal strength of K562-specific ChIP-seq peak regions defined by (A) SMC3, (B) ZNF143, (C) CTCF, and (D) RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.

**Supplementary Figure 7. Known architectural transcription factors were statistically significantly closer to predicted TAD boundaries then called TAD boundaries.** Barplots illustrate the log2 distance in basepairs from classes of TAD boundaries to transcription factors including CTCF, RAD21, SMC3, and ZNF143 for the GM12878 (A) and K562 (B) cell lines. TAD boundaries were classified into the 3 classes. Predicted but not called (PBNC) were the set of 10 kb flanked predicted TAD boundaries the did not overlap with 10 kb flanked called TAD boundaries. Called and predicted (CAP) were the set of 10 kb flanked called and predicted TAD boundaries that overlapped with each other. Called but not predicted (CBNP) were the set of 10 kb flanked called TAD boundaries that did not overlap with 10 kb flanked predicted TAD boundaries.
-->

>>>>

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we assessed the hypothesis that TADs and chromatin loops are conserved  [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa] by comparing predicted and called boundaries between the GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

@@

![**preciseTAD boundaries are enriched in CTCF sites.** (A) Visualizing locations of predicted vs called ARROWHEAD TAD boundaries on GM12878 (chr 1: 70600000-71530000). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. The blue vertical lines represent boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by our random forest. The green vertical lines represent the *preciseTAD boundary points* (PTBP) predicted by *preciseTAD*, with the corresponding demarcated TADs outlined above in green. Below is a zoomed in portion of the genome showing a cluster of base pairs with high probability representing a single *preciseTAD boundary region (PTBR) colored in yellow. (B) Enriched heatmaps comparing the enrichment of CTCF sites around 5 kb flanked preciseTAD and ARROWHEAD TAD boundaries respectively. (C) Enriched heatmaps comparing the enrichment of CTCF sites around 10 kb flanked preciseTAD and PEAKACHU loop boundaries respectively. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF site between preciseTAD boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively. (E) A atrix of pairwise Jaccard indices across resolutions shows more invariance toward resolution for *preciseTAD* (upper triangular) compared to ARROWHEAD (lower triangular).???update when ready](Figures/figure5_enrichment_and_log2dist_ctcf.png)

@@

@@

**Supplementary Table 6.** Additional performance metrics when implementing a random forest using ARROWHEAD ground truth TAD boundaries.

**Supplementary Table 7.** Additional performance metrics when implementing a random forest using PEAKACHU ground truth chromatin loop boundaries.
